\pdftitle

### 2018-05-15 maxNt_v3.tex
V3

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T13:47:43+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

% \DeclareMathSymbol{\nu}{\mathalpha}{egreeki}{"17}
% \DeclareMathSymbol{\xi}{\mathalpha}{egreeki}{"18}

@@

% \DeclareMathSymbol{\sigma}{\mathalpha}{egreeki}{"1B}

@@

% \DeclareMathSymbol{\varXi}{\mathalpha}{egreeki}{"04}

@@

% \DeclareMathSymbol{\varSigma}{\mathalpha}{egreeki}{"06}

@@

\newcommand*{\yxx}{x}%subpop state
\newcommand*{\yx}{\bm{\yxx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{X}}%pop state
\newcommand*{\yy}{\bm{y}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{r}}%subpop value
\newcommand*{\ys}{\bm{s}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{R}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varUpsilon}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
%Lagr multipliers
\newcommand*{\yf}{\bm{f}}
\newcommand*{\yc}{\bm{c}}
\newcommand*{\yL}{\bm{\varLambda}}
\newcommand*{\yl}{\bm{\lambda}}
\newcommand*{\yk}{\zeta}
\newcommand*{\yK}{\varZeta}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

\begin{abstract}
  This note shows that the \me\ method can be applied to a representative
  sample from a neuronal population along two different routes: (1) apply
  to the sample; or (2) apply to the population and marginalize to the
  sample. These two routes give inequivalent results. Which route should be
  chosen? Some arguments are presented in favour of the second. The note
  also touches upon probability formulae of representative sampling and
  discusses their possible meanings, a discussion that may be useful for
   sampling problems in neuroscience.
  % The abstract must be limited to one paragraph.
\end{abstract}

@@

Suppose that we have recorded the firing activity of a hundred neurons,
sampled from a particular brain area. This scenario is a concrete
possibility thanks to recent electrophysiological techniques
\cite{berenyietal2014}. In 1985, Gerstein, Perkel, \amp\ Dayhoff wrote (our
emphasis):
\begin{quote}
  the very profusion of data produced in such experiments creates serious
  problems of analysis and interpretation; these problems are conceptual as
  well as practical. The principal conceptual problems are (1)
  \emph{defining cooperativity or functional grouping} among neurons and
  (2) \emph{formulating quantitative criteria} for recognizing and
  characterizing such cooperativity. \cite{gersteinetal1985}
\end{quote}
The neuroscientific literature has since offered several mathematical
definitions of \enquote{cooperativity} or \enquote{functional grouping} and
criteria to quantify it.

One such quantitative criterion relies on the \me\ method
\cite{jaynes1957,jaynes1963,hobsonetal1973,sivia1996_r2006,meadetal1984}.
This criterion has been used in neuroscience at least since the 1990s,
applied to data recorded from brain areas as diverse as retina and motor
cortex
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015},
and it has been subjected to mathematical and conceptual scrutiny
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.

\enquote{Cooperativity} can quantified and characterized with \me\ method
in several ways. The simplest way proceeds along the following steps.
Consider the recorded activity of a sample of $n$ neurons.
\begin{enumerate}
\item The activity of each neuron, ideally a continuous signal, is
  transformed into a sequence of \enquote{$0$}s (inactive) and
  \enquote{$1$}s (firing) by binning the time and binarizing the signal
  intensity \cite{caianiello1961,caianiello1986}.

\item The total activity $n\yxs$ of the sample in each time bin is
  therefore an integer between $0$ and $n$. The total pairwise activity in
  each time bin, \ie\ the total number of simultaneously active pairs, is
  combinatorially given by $\tbinom{n\yxs}{2}\equiv n\yxs\,(n\yxs-1)/2$,
  and so on for higher-order simultaneous activities.

\item We construct a collection of \me\ distribution for the total
  activity, using as constraints: the time-averaged total activity; the
  time-averaged total activity \emph{and} total pairwise activity; and so
  on up to the time-averaged total activities of orders $1$ to $k$. Call
  the resulting distributions $p_1(n\yxs), p_2(n\yxs),\dotsc,p_k(n\yxs)$.
  We also have the empirical frequency distribution of the total activity,
  $f(n\yxs)$, counted from the time bins.

\item We can now compare the distributions above with one another and with
  the frequency distribution, using some probability-space distance like
  the relative entropy or discrimination information
  \citep{kullback1987,jaynes1963,hobson1969,hobsonetal1973}. If we find,
  for example, that such distance is very high between $p_1$ and $f$, but
  very low between $p_k$ and $f$ for $k \ge 2$, then we can say that there
  is a \enquote{pairwise cooperativity}, and that any higher-order
  cooperativity is just a reflection or consequence of the pairwise one.
\end{enumerate}
The protocol above needs to be made precise by specifying various
parameters, for example the width of the time bins.

We hurry to say that the description just given is just \emph{one} way to
quantify and characterize cooperativity and functional grouping, not
\emph{the} only way. It can surely be criticized from many points of view.
Yet, it is quantitative and more precise than an undefined notion of
\enquote{cooperativity}. Two people who apply it to the same data will
obtain the same numbers. Other protocols can be based on the \me\ method,
for example protocols that take into account the activities or joint
activities of individual neurons, or even protocols that take into account
time dependence.

The purpose of the present work is to show that there is a problem in the
way the \me\ method as sketched above, or more complex variations of it, is
applied to the activity of the recorded neurons. The problem stems from the
fact that the recorded neurons are a sample from a larger population.

Suppose that the neurons were recorded with electrodes covering an area of
some square millimetres. This recording is a sample of the activity of the
population covered by the electrode; we can consider it as a sample of an
even more extended brain area. The characterization of the cooperativity of
the recorded sample would have little meaning if we did not expect its
results to generalize to a larger area, at least that covered by the
electrodes. In other words, we expect that the conclusions drawn with the
\me\ methods about the sampled neurons should somehow generalize or
extrapolate at least to larger areas from which those neurons were sampled.
We are thus assuming that the recorded sample is a \emph{representative
  sample} of some larger area. Probability theory tells us how to make
inferences from a sample to the larger population from which it is sampled
\parentext{see references below}.

Here is the problem. We can apply the \me\ method to the sample, as
described in the above protocol, to generate probability distributions for
the state of the sample. But, if our sample is representative of a larger
population, we can apply the \me\ method to the larger population, using
the constraints obtained from the sample, and generate a probability
distribution for the larger population; the distribution for the sample is
then obtained by marginalization. Either application seems to have some
commendable features. However, \emph{the distributions obtained from these
  two applications differ}. Which choice is most meaningful?

@@

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$; the \enquote{$X$} denote the
variables, and \enquote{$R$} their values. A particular sample of $n$ neurons
from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$ having
fixed binary values $(r_1, \dotsc, r_n)\defs\yr \in \set{0,1}^n$. We will
consider various averages of the population and the sample. For this
purpose we introduce a population-averaging operator $\sav{\dotv}$ defined
by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    \qquad\text{and so on.}
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

@@

Our uncertainty about the actual state of the population is completely
expressed by the joint probability distribution
\begin{equation}
  \label{eq:joint_plaus}
  \p(X_1=R_1, X_2=R_2, \dotsc, X_N=R_N \| \yHc) \quad\text{or}\quad
\p(\yX =\yR \| \yHc), \quad \yR \in \set{0,1}^N,
\end{equation}
where $\yHc$ denotes our state of knowledge, \ie\ the evidence and
assumptions backing this particular probability assignment. Our uncertainty
about the state of the sample is likewise expressed by
\begin{equation}
  \label{eq:sample_plaus}
  \p(x_1=r_1, x_2=r_2, \dotsc, x_n=r_n \| \yHc) \quad\text{or}\quad
\p(\yx =\yr \| \yHc), \quad \yr \in \set{0,1}^n.
\end{equation}

@@

The \emph{representation theorem for finite exchangeability} states that
the symmetric distribution in the state of knowledge $\yH$ must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \|  \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, \yH)\,
  \p(\yXf=\yRf \| \yH) =
  \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. A sum is
only apparently missing in the central term: its summands $\yXf=A$ are all
zero except for $A=\yRf$. Proof of this theorem and generalizations to
non-binary and continuum cases are given by de~Finetti
\cite{definetti1959b}, Kendall \cite{kendall1967}, Ericson
\cite{ericson1976}, Diaconis \amp\ Freedman
\cite{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\cite{heathetal1976}. This theorem is intuitive: owing to symmetry, we must
assign equal probabilities to all states with $N\yRf$ active neurons.

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
The conditional probability in the last formula is a hypergeometric
distribution $\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of the formulae above is
in fact the same as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
Our initial symmetric ignorance should intuitively also apply to the
sample; the probability for the state of the
sample~\eqref{eq:subpop_average} indeed satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

The conditional probability
$\p(\yxs = \yrs \|\yXf=\yRf, \yH) \equiv \ypp(\yrs,\yRf)$ relates the
spaces of the sample average $\yXf\in\set{0,\dotsc,N}$ and of the
population average $\yxs\in\set{0,\dotsc,n}$. It is a coarsening projector
of any probability for $\yXf$ onto a marginal probability for $\yxs$, as
\eqn~\eqref{eq:subpop_average} shows. For each function $f$ of the sample
average $\yxs$, the  hypergeometric distribution 
$\ypp(\yrs,\yRf)$ also allows us to construct a function $f^*$ of the
population average $\yXf$
\begin{equation}
  \label{eq:pullback_P}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
\end{equation}
that has the same expectation as $f$:
\begin{multline}
  \expeb{f^*(\yXf)}
  \equiv
  \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH)
  =
  \sum_{N\yRf=0}^N \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yRf)\,\p(\yXf=\yRf \| \yH)
  ={}\\
  \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
  \equiv  \expeb{f(\yxs)}
\end{multline}

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T22:50:14+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

 \DeclareMathSymbol{\nu}{\mathalpha}{egreeki}{"17}
 \DeclareMathSymbol{\xi}{\mathalpha}{egreeki}{"18}

@@

 \DeclareMathSymbol{\sigma}{\mathalpha}{egreeki}{"1B}

@@

 \DeclareMathSymbol{\varXi}{\mathalpha}{egreeki}{"04}

@@

 \DeclareMathSymbol{\varSigma}{\mathalpha}{egreeki}{"06}

@@

\newcommand*{\yvX}{\varSigma}
\newcommand*{\yvR}{\sigma}
\newcommand*{\yvx}{S}
\newcommand*{\yvr}{s}
\newcommand*{\yvN}{\nu}

\newcommand*{\yx}{\bm{\yvx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{\yvX}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{\yvr}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{\yvR}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varUpsilon}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
%Lagr multipliers
\newcommand*{\yf}{\bm{f}}
\newcommand*{\yc}{\bm{c}}
\newcommand*{\yL}{\bm{\varLambda}}
\newcommand*{\yl}{\bm{\lambda}}
\newcommand*{\yk}{\zeta}
\newcommand*{\yK}{\varZeta}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

\begin{abstract}
  \mynote{to be rewritten} This note shows that the \me\ method can be
  applied to a representative sample from a neuronal population along two
  different routes: (1) apply to the sample; or (2) apply to the population
  and marginalize to the sample. These two routes give inequivalent
  results. Which route should be chosen? Some arguments are presented in
  favour of the second. The note also touches upon probability formulae of
  representative sampling and discusses their possible meanings, a
  discussion that may be useful for sampling problems in neuroscience.
  % The abstract must be limited to one paragraph.
\end{abstract}

@@

Suppose that we have recorded the firing activity of a hundred neurons,
sampled from a particular brain area. What are we to do with such data?
Gerstein, Perkel, \amp\ Dayhoff \cite{gersteinetal1985} posed this question
very tersely (our emphasis):
\begin{quote}
  The principal conceptual problems are (1) \emph{defining cooperativity or
    functional grouping} among neurons and (2) \emph{formulating
    quantitative criteria} for recognizing and characterizing such
  cooperativity.
\end{quote}
These questions have a long history, of course; see \eg\ the 1966 review by
Moore \etal\ \cite{mooreetal1966}. % This scenario is a concrete possibility
% thanks to recent electrophysiological techniques \cite{berenyietal2014}.
The neuroscientific literature has offered several mathematical definitions
of \enquote{cooperativity} or \enquote{functional grouping} and criteria to
quantify it.

One such quantitative criterion relies on the \me\ or relative-\me\ method
\cite{jaynes1957,jaynes1963,hobsonetal1973,sivia1996_r2006,meadetal1984}.
This criterion has been used in neuroscience at least since the 1990s,
applied to data recorded from brain areas as diverse as retina and motor
cortex
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015},
and it has been subjected to mathematical and conceptual scrutiny
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.

\enquote{Cooperativity} can quantified and characterized with \me\ method
in several ways. The simplest way roughly proceeds along the following
steps. Consider the recorded activity of a sample of $n$ neurons.
\begin{enumerate}
\item The activity of each neuron, a continuous signal, is binned in time
  and binarized in intensity, and thus transformed into a sequence of
  digits \enquote{$0$}s (inactive) and \enquote{$1$}s
  \cite[\cf][]{caianiello1961,caianiello1986}.

  Let the variable $\yvx_i(t)\in\set{0,1}$ denote the activity of the $i$th
  sampled neuron at time bin $t$. The average activity at that bin is
  therefore $\yxs(t) \defd \sum_i \yvx_i(t)/n$. If we count the number of
  distinct pairs of active neurons at that bin we combinatorially find
  $\tbinom{n\yxs(t)}{2}\equiv n\yxs(t)\,[n\yxs(t)-1]/2$. There can be at
  most $\tbinom{n}{2}$ simultaneously active pairs, so the average pair
  activity is
  $\av{\yx \yx}(t) \defd \tbinom{n}{2}^{-1}\tbinom{n\yxs(t)}{2}$. With some
  combinatorics we see that the average activity of $m$-tuples of neurons
  is
  \begin{equation}
    \label{eq:products_intermsof_average}
    \av{\underbrace{\yx\dotsm\yx}_{\text{$m$ terms}}}
    = \binom{n}{m}^{-1}\binom{n\yxs(t)}{m}.
  \end{equation}
  For brevity let us agree to simply call \enquote{activity} the average
  $\yxs$, \enquote{pair-activity} the average $\av{\yx \yx}$, and so on.

\item Construct a sequence of relative-\me\ distributions for the activity
  $\yxs$, using this sequence of constraints:
  \begin{itemize}
  \item the time average of the activities $\yxs(t)$;
  \item the time averages of the activities $\yxs(t)$ and pair-activities
    $\av{\yx \yx}(t)$;
  \item \ldots and so on, including the time averages of $\yxs(t)$ up to
    $\av{\yx\dotsm\yx}(t)$ with $k$ terms.
  \end{itemize}
  Call the resulting distributions $p_1(\yxs), p_2(\yxs),\dotsc,p_k(\yxs)$.
  The time-bin dependence is now absent because these distributions can be
  interpreted as referring to any one of the time bins $t$, or to a new
  time bin containing new data.

  We also have the empirical frequency distribution of the total activity,
  $f(\yxs)$, counted from the time bins.

\item Now compare the distributions above with one another and with the
  frequency distribution, using some probability-space distance like the
  relative entropy or discrimination information
  \citep{kullback1987,jaynes1963,hobson1969,hobsonetal1973}. If we find,
  say, that such distance is very high between $p_1$ and $f$, very low
  between $p_2$ and $f$, and is more or less the same between all $p_m$ and
  $f$ for $m \ge 2$, then we can say that there is a \enquote{pairwise
    cooperativity}, and that any higher-order cooperativity is just a
  reflection or consequence of the pairwise one. The reason is that the
  information from higher-order simultaneous activities did not lead to
  appreciable changes in the distribution obtained from pair activities.
\end{enumerate}
The protocol above needs to be made precise by specifying various
parameters, such as the width of the time bins or the probability
distance used.

We hurry to say that the description just given is just \emph{one} way to
quantify and characterize cooperativity and functional grouping, not
\emph{the only} way. It can surely be criticized from many points of view.
Yet, it is quantitative and bears a more precise meaning than an undefined,
vague notion of \enquote{cooperativity}. Two persons who apply this
procedure to the same data will obtain the same numbers. Different
protocols can be based on the \me\ method, for instance protocols that take
into account the activities or pair activities of specific neurons rather
than population averages, or even protocols that take into account time
dependence.

The purpose of the present work is not to assess the merits of \me\ methods
with respect to other methods. Its main purpose is to show that there is a
problem in the way the \me\ method itself, as sketched above, is applied to
the activity of the recorded neurons. We believe that this problem is at
the root of some quirks about this method that were pointed out in the
literature \cite{roudietal2009b}. This problems extends also to more
complex versions of the method, possibly except version that use
\enquote{hidden} neurons. The problem stems from the fact that the recorded
neurons are a \emph{sample} from a larger, unrecorded population.

Suppose that the neurons were recorded with electrodes covering an area of
some square millimetres \cite[\cf][]{berenyietal2014}. This recording is a
sample of the activity of the neuronal population under the recording
device; such population can amount to tens of thousands of neurons
\cite{abeles1991}. We could even consider the recorded neurons as a sample
of a brain area more extended than the recording device.

The characterization of the cooperativity of the recorded sample would have
little meaning if we did not expect its results to generalize to a larger,
unrecorded population -- at the very least that under the recording device.
In other words, we expect that the conclusions drawn with the \me\ methods
about the sampled neurons should somehow extrapolate to unrecorded neurons
in some larger area, from which the recorded neurons were sampled. In
statistical terms we are assuming that the recorded neurons are a
\emph{representative sample} of some larger neuronal population.
Probability theory tells us how to make inferences from a sample to the
larger population from which it is sampled \parentext{see references
  below}.

Here is the problem. We can apply the \me\ method to the sample, as
described in the above protocol, to generate probability distributions for
the activity of the sample. But, if our sample is representative of a
larger population, we can apply the \me\ method to the larger (unrecorded)
population, using the constraints obtained from the recorded sample, and
generate a probability distribution for the larger population. The
distribution for the sample is then obtained by marginalization. Either
application seems to have some commendable features. However, \emph{the
  distributions obtained from these two applications differ}. Which choice
is most meaningful?

@@

We have a population of $\yvN$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(\yvX_1,\dotsc,\yvX_\yvN) \defs \yX$ having fixed but unknown binary values
$(\yvR_1,\dotsc,\yvR_\yvN)\defs \yR \in \set{0,1}^\yvN$; the \enquote{$\yvX$} denote the
variables, and \enquote{$\yvR$} their values. A particular sample of $n$ neurons
from this population has joint state $(\yvx_1, \dotsc, \yvx_n) \defs \yx$ having
fixed binary values $(\yvr_1, \dotsc, \yvr_n)\defs\yr \in \set{0,1}^n$. We will
consider various averages of the population and the sample. For this
purpose we introduce a population-averaging operator $\sav{\dotv}$ defined
by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{\yvN} (\yvX_1 + \yvX_2 + \dotsb + \yvX_\yvN),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{\yvN}{2}^{-1} (\yvX_1 \yvX_2 + \yvX_1 \yvX_3 + \dotsb +
    \yvX_{\yvN-1} \yvX_\yvN),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{\yvN}{3}^{-1} (\yvX_1 \yvX_2 \yvX_3 +% \yvX_1 \yvX_2 \yvX_4
    \dotsb + \yvX_{\yvN-2} \yvX_{\yvN-1}\yvX_\yvN),
    \qquad\text{and so on.}
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{\yvN}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $\yvX_i \dotsm \yvX_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{\yvN}{m}^{-1}\binom{\yvN\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

@@

Our uncertainty about the actual state of the population is completely
expressed by the joint probability distribution
\begin{equation}
  \label{eq:joint_plaus}
  \p(\yvX_1=\yvR_1, \yvX_2=\yvR_2, \dotsc, \yvX_\yvN=\yvR_\yvN \| \yHc) \quad\text{or}\quad
\p(\yX =\yR \| \yHc), \quad \yR \in \set{0,1}^\yvN,
\end{equation}
where $\yHc$ denotes our state of knowledge, \ie\ the evidence and
assumptions backing this particular probability assignment. Our uncertainty
about the state of the sample is likewise expressed by
\begin{equation}
  \label{eq:sample_plaus}
  \p(\yvx_1=\yvr_1, \yvx_2=\yvr_2, \dotsc, \yvx_n=\yvr_n \| \yHc) \quad\text{or}\quad
\p(\yx =\yr \| \yHc), \quad \yr \in \set{0,1}^n.
\end{equation}

@@

The \emph{representation theorem for finite exchangeability} states that
the symmetric distribution in the state of knowledge $\yH$ must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \|  \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, \yH)\,
  \p(\yXf=\yRf \| \yH) =
  \binom{\yvN}{\yvN\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. A sum is
only apparently missing in the central term: its summands $\yXf=A$ are all
zero except for $A=\yRf$. Proof of this theorem and generalizations to
non-binary and continuum cases are given by de~Finetti
\cite{definetti1959b}, Kendall \cite{kendall1967}, Ericson
\cite{ericson1976}, Diaconis \amp\ Freedman
\cite{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\cite{heathetal1976}. This theorem is intuitive: owing to symmetry, we must
assign equal probabilities to all states with $\yvN\yRf$ active neurons.

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{\yvN \yRf =0}^{\yvN}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{\yvN-n}{\yvN \yRf-n\yrs}\binom{\yvN}{\yvN \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
The conditional probability in the last formula is a hypergeometric
distribution $\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of the formulae above is
in fact the same as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
Our initial symmetric ignorance should intuitively also apply to the
sample; the probability for the state of the
sample~\eqref{eq:subpop_average} indeed satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

The conditional probability
$\p(\yxs = \yrs \|\yXf=\yRf, \yH) \equiv \ypp(\yrs,\yRf)$ relates the
spaces of the sample average $\yXf\in\set{0,\dotsc,\yvN}$ and of the
population average $\yxs\in\set{0,\dotsc,n}$. It is a coarsening projector
of any probability for $\yXf$ onto a marginal probability for $\yxs$, as
\eqn~\eqref{eq:subpop_average} shows. For each function $f$ of the sample
average $\yxs$, the  hypergeometric distribution 
$\ypp(\yrs,\yRf)$ also allows us to construct a function $f^*$ of the
population average $\yXf$
\begin{equation}
  \label{eq:pullback_P}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
\end{equation}
that has the same expectation as $f$:
\begin{equation}
  \expeb{f^*(\yXf)}
  \equiv
  \sum_{\yvN\yRf=0}^\yvN f^*(\yRf)\,\p(\yXf=\yRf \| \yH)
%  =
%  \sum_{N\yRf=0}^N \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yRf)\,\p(\yXf=\yRf \| \yH)
  =
  \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
  \equiv  \expeb{f(\yxs)}.
\end{equation}
The middle equality uses the combination of formulae~\eqref{eq:pullback_P}
and~\eqref{eq:subpop_average}.

\begin{figure}[!t]
\centering
\includegraphics[width=0.9\linewidth]{pop_sample_projection3.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
\ypp(\yrs\|\yRf) \defd  \raisebox{0pt}[0pt][1ex]{}\binom{n}{n\yrs}\binom{\yvN-n}{\yvN \yRf-n\yrs}\binom{\yvN}{\yvN \yRf}^{-1}$ for $\yvN=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, see
\fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of \enquote{fuzzy
  identity matrix} between the $\yXf$-space $\set{0,\dotsc,\yvN}$ and
$\yxs$-space $\set{0,\dotsc,n}$. When $n=\yvN$ it is the identity matrix. We
thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities: they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.
